---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Research topic: Hotel Booking Data 

### Team members: Kaitlyn Hoyme and Christopher Moseley  

### Introduction:  

This report will demonstrate various relationships between variables surrounding hotel booking data. We will examine these variables in order to determine how they affect each other and hotel bookings. This data would be of interest to businesses like hotels and travel agencies, as they can adjust these factors to improve their profit and customer satisfaction. It could also be useful for customers to observe what factors might make their hotel bookings cheaper or more satisfactory.   

  
### Questions we will address: 

- How does the time of year affect hotel bookings?  
- What makes guests more likely to cancel their bookings?  
- How does the average daily rate affect bookings?  
- How does number of guests affect hotel bookings?  
- What affects how many times bookings are changed?  
- Where are most visitors coming from?   

  
By investigating these questions we hope to demonstrate useful data about what factors affect hotel bookings.    
  

### Data:   

This dataset includes information on hotel booking data from two hotels, one a city hotel, the other a resort hotel, with data collected between the 1st of July 2015 and 31st of August 2017.   

There are 119390 rows of 36 variables.  
The Variables are:   
 - hotel : States what hotel the row represents data from. City Hotel or Resort Hotel.  
 - is_canceled : value that indicates if the booking was canceled (1) or not (0).  
 - lead_time : Number of days that occurred between the entering date of the booking into the PMS and the arrival date.  
 - arrival_date_year : The year of the arrival date (values are between 2015-2017).  
 - arrival_date_month : Month of arrival date. Categorical variable with values "January" through "December".  
 - arrival_date_week_number : Week number of the arrival date.  
 - arrival_date_day_of_month : Day of the month of the arrival date.  
 - stays_in_weekend_nights : Number of weekend nights (Saturday or Sunday) the guest stayed or booked to stay at the hotel.   
 - stays_in_week_nights :  Number of week nights (Monday-Friday) the guest stayed or booked to stay at the hotel.   
 - adults : Number of adults.  
 - children : Number of children.  
 - babies : Number of babies.  
 - meal : Represents the type of meals offered by the hotel. Ex: BB = Bed & Breakfast.  
 - country : Country of origin.  
 - market_segment : Market segment designation. In categories (categorical). "TA" means travel agents, and "TO" means tour operators    
 - distribution_channel : Booking distribution channel. The term "TA" means travel agents, and "TO" means tour operators.  
 - is_repeated_guest : Value indicating if the booking name was from a repeated guest (1) or not (0).  
 - previous_cancellations : Number of previous bookings that were cancelled by the customer prior to the current booking.   
 - previous_bookings_not_canceled : Number of previous bookings not cancelled by the customer prior to the current booking.   
 - reserved_room_type : Code of room type reserved.   
 - assigned_room_type : Code for type of room assigned to the booking. Sometimes the assigned room type differs from the reserved room type due to hotel operation reasons (eg overbooking) or by customer request.     
 - booking_changes : Number of changes/amendments made to the booking from the moment the booking was entered on the PMS until the moment of check-in or cancellation.   
 - deposit_type : Represents deposit type. No Deposit - no deposit was made; Non Refund - a deposit was made in the value of the total stay cost; Refundable - a deposit was made with a value under the total cost of stay.   
 - agent : ID of the travel agency that made the booking.  
 - company : ID of the company/entity that made the booking or responsible for paying the booking. ID is presented instead of designation for anonymity reasons.    
 - days_in_waiting_list : Number of days the booking was in the waiting list before it was confirmed to the customer.   
 - customer_type : Group - when the booking is associated to a group; Transient - when the booking is not part of a group or contract and is not associated to other transient booking; Transient-party - when the booking is transient but is associated to at least one other transient booking.   
 - adr : Average daily rate calculated by dividing the sum of all lodging transactions by the total number of staying nights.  
 - required_car_parking_spaces : Number of parking spaces required by the customer.   
 - total_of_special_requests : Number of special requests made by the customer (e.g. twin bed or high floor).  
 - reservation_status : Check-Out - customer has checked in but already departed; No-Show - customer did not check-in and did inform the hotel of the reason why, etc..  
 - reservation_status_date : Date at which the last status was set. This variable can be used in conjunction with the ReservationStatus to understand when the booking was canceled or when the customer check-out of the hotel.  
 - name : Name of the guest (not real for anonymity).  
 - email : Email of the guest (not real for anonymity).  
 - phone-number : Phone number of the guest (not real for anonymity).  
 - credit_card : Credit Card Number of the guest (not real for anonymity).    
 
This dataset was found through Kaggle and all of the data recorded is real. However, for reasons surrounding anonymity variables like name, email, phone-number, and credit card are all made up. The data was originally obtained from "Hotel Booking Demand Datasets", an article written by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.   
This is a link to our dataset on Kaggle: https://www.kaggle.com/datasets/mojtaba142/hotel-booking/data  

### Data Cleaning
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
#Data: description of your data set, first data cleaning steps, marginal summaries;

url <- "https://raw.githubusercontent.com/khoy24/DS202FinalProject/main/hotel_booking.csv"
data <- read.csv(url)
colnames(data)


#nrow(data) # 119390 rows/entries
#ncol(data) # 36 variables

# any(is.na(data))
# a few missing entries
#colSums(is.na(data))
#only missing columns were in agent and company which makes sense so we decided to leave that. There are also
#4 NA values in the children column which we will leave. 

duplicateRows <- data %>% filter(duplicated(.))
# print(nrow(duplicateRows))
# no duplicate rows in this dataset

#examine types of columns, make sure makes sense
str(data)

#Change variable types to be correct
#Need to make year not be an int. Change that and the similar values to strings. Also had to make is_repeated_guest a character so it wasn't treated numerically when graphing.
data <- data %>%
  mutate(
    `arrival_date_year` = as.character(`arrival_date_year`),
    `arrival_date_week_number` = as.character(`arrival_date_week_number`),
    `arrival_date_day_of_month` = as.character(`arrival_date_day_of_month`),
    `is_repeated_guest` = as.character(`is_repeated_guest`)
  )


```

### Results  

### How does the time of year affect hotel bookings?  
```{r, message=FALSE, warning=FALSE}

#Have to make months a factor to reorder in order of correct months
data$arrival_date_month <- factor(data$arrival_date_month, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

# We also only have data for half the year of 2015, and 8 months of 2017. 2016 was the only full year we had. This is why we represent it by months, not years. 

#We separate hotel bookings per month and facet that by year, as we'd have more months for those included on the half years. 

#color code the months by temperature 
monthcolors <- c(
  "January" = "#4575b4",  
  "February" = "#74add1",  
  "March" = "#abd9e9",
  "April" = "#e0f3f8",
  "May" = "#fee090",
  "June" = "#fdae61",
  "July" = "#f46d43",     
  "August" = "#d73027",   
  "September" = "#fdae61",
  "October" = "#fee090",
  "November" = "#abd9e9",
  "December" = "#74add1"
)

ggplot(data, aes(x = arrival_date_month, fill = arrival_date_month)) +
  geom_bar(width = 0.8) + 
  labs(title = "Hotel Bookings per Month", 
       x = "Month", y = "Number of Bookings") +
  facet_wrap(~ arrival_date_year) +
  scale_fill_manual(values = monthcolors) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  )
```
  
This plot shows the number of hotel bookings per month (it shows the number of bookings made FOR that month, not what month the booking was made).  

Now we see that in 2015 the months with the most bookings were September-October, in 2016 it was October again, followed by the spring and summer months, and in 2017 it was May with the highest number of bookings, with its surrounding months following it. For the most part we still see that months with warmer. more mild weather, like the spring and fall seasons have more bookings.  

Now we observe how how the month affects the bookings separately for the resort and city hotels.  
```{r, message=FALSE, warning=FALSE}

ggplot(data, aes(x = arrival_date_month, fill = hotel)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ arrival_date_year) +
  labs(title = "Monthly Bookings: Resort vs City Hotel", x = "Month", y = "Bookings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))


```
  
For the most part the two hotels follow relatively the same trend with the months with a few exceptions. We also notice that the city hotel has a lot more bookings in comparison to the resort hotel. The city hotel also has more variation in the number of bookings, while the resort hotels number of bookings seems more consistent.  

The following graph shows numbers of bookings by month and year in heat map form.  
```{r, message=FALSE, warning=FALSE}

heatmap_data <- data %>%
  group_by(arrival_date_year, arrival_date_month) %>%
  summarise(bookings = n())

ggplot(heatmap_data, aes(x = arrival_date_month, y = arrival_date_year, fill = bookings)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#e0f3f8", high = "#d73027") +
  labs(title = "Hotel Bookings by Month and Year",
       x = "Month", y = "Year", fill = "Bookings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

```
  
As you can see in this graph, we can see that the bookings in generally have been increasing slightly over the years, and that they are generally higher in the spring and the fall, as confirmed in our earlier graph. However, this gives us a better idea of how the trends are changing over the years.    

The following graph shows bookings cancelled vs bookings not cancelled by month and facetted by year.  
```{r, message=FALSE, warning=FALSE}

# set up data to compare with cancelled or not
cancellation_summary <- data %>%
  group_by(arrival_date_year, arrival_date_month, is_canceled) %>%
  summarise(bookings = n(), .groups = 'drop')

# add key
cancellation_summary$is_canceled <- factor(cancellation_summary$is_canceled,
                                           labels = c("Not Canceled", "Canceled"))

ggplot(cancellation_summary, aes(x = arrival_date_month, y = bookings, fill = is_canceled)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ arrival_date_year) +
  labs(title = "Hotel Booking Cancellations by Month and Year",
       x = "Arrival Month", y = "Number of Bookings", fill = "Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

```
  
Looking at the chart the number of cancellations seems to follow the number of bookings. There doesn't seem to be any super significant months or years where cancellations are more frequent in comparison to the number of bookings for that month. We found this a bit surprising as we thought winter months would have more cancellations to due unforeseen weather conditions, but we also do not know the weather patterns of the hotel locations so it is hard to say.  

Now we want to observe if the month/year affects the duration of a guest's stay.  

```{r, message=FALSE, warning=FALSE}

#add variables for length of stay for the graph
data <- data %>%
  mutate(
    length_of_stay = stays_in_weekend_nights + stays_in_week_nights,
  )

avg_stay_per_booking <- data %>%
  group_by(arrival_date_year, arrival_date_month) %>%
  summarise(
    avg_stay = mean(length_of_stay, na.rm = TRUE),
    .groups = 'drop'
  )


ggplot(avg_stay_per_booking, aes(x = arrival_date_month, y = avg_stay, fill = arrival_date_month)) +
  geom_bar(stat = "identity") +
  facet_wrap(~arrival_date_year) +
  labs(title = "Average Length of Stay per Booking",
       x = "Month", y = "Average Nights Stayed") +
  scale_fill_manual(values = monthcolors) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  )

```
  
This graph shows the average length of stay (in nights) for each booking. We calculate total nights stayed by adding weekend nights and week nights together. It demonstrates that the average length of stay for a booking is longer in summer months / hotter weather with it being around 3-4 nights, with the colder months being more around 2-3 nights on average.  

Overall, the important things to take away from these graphs are that the most bookings are made in the spring and fall months, with longer stays the warmer the weather is (summer months), the city hotel has more varying amounts of bookings made throughout the year in comparison to the resort hotel, cancellation rates don't seem to be dramatically affected by season, and that as the years have gone by there have been more bookings made.  

### What makes guests more likely to cancel their bookings?  

We will examine the following variables for how they affect cancellations:  
- lead_time  
- previous_cancellations  
- returning guest  
- customer_type  
- market_segment  
- hotel  
- special_requests  
  
The following graph shows cancellation by lead time.  
```{r, message=FALSE, warning=FALSE}

ggplot(data, aes(x = lead_time, fill = factor(is_canceled))) +
  geom_histogram(position = "fill", bins = 50) +
  labs(title = "Cancellation Rate by Lead Time", x = "Lead Time (days)",
       y = "Proportion", fill = "Canceled") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5)
  )

```
  
Apart from the two outliers on the far right of the graph that did not cancel their bookings, the trend seems to be that the longer the lead time the higher the chance of cancellation. This makes sense with what we would have assumed as when you book closer to the arrival date you have a better idea of what you need. In contrast, when you book farther out from the date things are more likely to change.    

The following graph demonstrates how the number of previous cancellations affects the chance of cancelling again.  
```{r, message=FALSE, warning=FALSE}

data %>%
  group_by(previous_cancellations) %>%
  summarise(
    count = n(),
    cancel_rate = mean(is_canceled),
    .groups = "drop"
  ) %>%
  filter(count > 20) %>%  # ignore values with less data entries for viewing purposes, it was too large before
  ggplot(aes(x = as.factor(previous_cancellations), y = cancel_rate, fill = previous_cancellations)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = scales::percent(cancel_rate, accuracy = 0.1)), 
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Cancellation Rate by Number of Previous Cancellations",
    x = "Number of Previous Cancellations",
    y = "Cancellation Rate"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5)
  )

```
  
The graph demonstrates that once a person has 1 previous cancellation they are very likely to cancel again, then after that and they have 2 or more prior cancellations it drops to around 30% chance of cancellation, and goes up to 100% once it gets to around 24 previous cancellations. For example, 94.4% of customers that have previously cancelled once will cancel again on another booking. This graph shows that the rate of cancellation is generally lower for guests that have previously stayed there and didn't cancel at least once. While we expected the 0 previous cancellations to be around where it is for the cancellation rate, we assumed it would continually increase as the number of previous cancellations went up, but instead it spikes at 1 previous cancellation then stays relatively consistent until it gets to very large numbers of cancellations.  

   
Here we examine whether the fact that a guest is a returning guest or not affects the cancellation rate.  

```{r, message=FALSE, warning=FALSE}

ggplot(data, aes(x = factor(is_repeated_guest), y = is_canceled, fill = is_repeated_guest)) +
  stat_summary(fun = mean, geom = "bar", width = 0.6) +
  scale_x_discrete(labels = c("New Guest", "Returning Guest")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Cancellation Rate by Returning Guest Status",
    x = "Guest Status",
    y = "Cancellation Rate"
  ) +
  theme_minimal() +
  theme(
        legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )


```
  
As we can see, it seems as if new guests are more likely to cancel their bookings than a returning guest. This could be because returning guests know what they're getting when they book the hotel since they have booked it before and if they were returning knew they were going to be satisfied. It could also have something to do with guests returning frequently for things such as business trips, which then they will be more frequent and consistent customers.  
  
  
The graph shows the relationships from customer type to cancellations.  
```{r, message=FALSE, warning=FALSE}

data %>%
  group_by(customer_type) %>%
  summarise(
    total_bookings = n(),
    cancel_rate = mean(is_canceled),
    .groups = "drop"
  ) %>%
  mutate(customer_type = reorder(customer_type, cancel_rate)) %>%  # reorder to appear in increasing order
  ggplot(aes(x = customer_type, y = cancel_rate, fill = customer_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = percent(cancel_rate, accuracy = 0.1)), 
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cancellation Rate by Customer Type",
    x = "Customer Type",
    y = "Cancellation Rate"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )


```
  
This graph demonstrates the cancellation rates for various customer types. As we can see, transient customers have the highest cancellation rates with 40.7% cancellation rate. Group customers have the lowest with a 10.2% cancellation rate. This follows what we would expect, as groups would have things planned in advance and be more strict about sticking to plans, while transient individuals have more freedom with their bookings and probably more flexibility to change their schedules and plans.     


This graph demonstrates the relationship between market segments and cancellation rates. 
```{r, message=FALSE, warning=FALSE}

data %>%
  filter(market_segment != "Undefined") %>%
  group_by(market_segment) %>%
  summarise(
    total_bookings = n(),
    cancel_rate = mean(is_canceled),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(market_segment, -cancel_rate), y = cancel_rate, fill = market_segment)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = percent(cancel_rate, accuracy = 0.1)),
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cancellation Rate by Market Segment",
    x = "Market Segment",
    y = "Cancellation Rate"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )

```
  
This graph shows the cancellation rate by market segment. As you can see, the Groups category for market segment has the highest cancellation rate with 61.1%, and complementary has the lowest with 13.1%. This result may seem suprising since the preceding graph had groups as the lowest cancellation rate, however, upon doing further research we outline what the difference is between these designations.  
In the terms of customer type, group represents an actual official group booking like coworkers, conference-goers, teams, etc., while market segment group includes groups of individuals as well as official groups such as those mentioned before. So, market segment 'group' probably includes some group, transient and transient party groups as well that don't fit into the other market segments. After realizing this, we realized the data did make sense, as the other market segments such as corporate or complementary have more incentive to go and more restrictions that would not allow them to cancel.   
  
This next graph shows the cancellation rates in the different hotels (resort vs city).  
```{r, message=FALSE, warning=FALSE}

data %>%
  group_by(hotel) %>%
  summarise(
    total_bookings = n(),
    cancel_rate = mean(is_canceled),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = hotel, y = cancel_rate, fill = hotel)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = percent(cancel_rate, accuracy = 0.1)),
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cancellation Rate by Hotel Type",
    x = "Hotel Type",
    y = "Cancellation Rate"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
    plot.title = element_text(hjust = 0.5)
    )


```
  
As shown in the graph, the city hotel has a higher cancellation rate of about 41.7% whereas the resort hotel had 27.8% cancellation rate. We assume this would be because more people at the resort hotel are vacationing so their trips are based off the destination, whereas maybe the city hotel would have more bookings related to work / other nearby hotels that would allow or require people to change their plans more. A resort hotel's location is generally the main attraction.     

This graph shows how the number of special requests correlates with the cancellation rate.  
```{r, message=FALSE, warning=FALSE}

data %>%
  group_by(total_of_special_requests) %>%
  summarise(
    count = n(),
    cancel_rate = mean(is_canceled),
    .groups = "drop"
  ) %>%
  filter(count > 30) %>%  # limit to data with more than 30 occurrences, takes out some columns to make it easier to read
  ggplot(aes(x = as.factor(total_of_special_requests), y = cancel_rate, fill=total_of_special_requests)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = percent(cancel_rate, accuracy = 0.1)), 
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Cancellation Rate by Number of Special Requests",
    x = "Number of Special Requests",
    y = "Cancellation Rate"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5)
  )


```
  
This graph shows that as the number of special requests made gets higher, the rate of cancellation becomes lower. We originally assumed the opposite as we would assume those customers would be more picky, but with more special requests it does make sense that perhaps the customers would be more decisive about where they are staying and after making all the requests to accommodate themselves, less likely to cancel and probably more satisfied with their stay.     

Overall, it seemed as though all of these variables somewhat affected cancellation rate. However, we found that the results of the comparison with the number of previous cancellations was somewhat unhelpful and inconclusive. There was no significant pattern that we think would benefit hotels in any significant way to know. But, we believe the other variables, lead time, returning guest status, customer type, market segment, hotel type, and number of special requests would be helpful variables to pay attention to when determining likeliness of booking cancellation.  
  
### How does the average daily rate affect bookings?
```{r, warning=FALSE, message=FALSE}

# get monthly ADR
monthly_adr <- data %>%
  group_by(arrival_date_year, arrival_date_month) %>%
  summarise(avg_adr = mean(adr, na.rm = TRUE),
            total_bookings = n(),
            cancellations = sum(is_canceled)) %>%
  ungroup()

# Find ADR by booking frequency
data %>%
  filter(adr > 0 & adr < 500) %>%
  mutate(adr_bin = cut(adr, breaks = seq(0, 500, by = 25))) %>%
  group_by(adr_bin) %>%
  summarise(bookings = n()) %>%
  ggplot(aes(x = adr_bin, y = bookings)) +
  geom_bar(stat = "identity") +
  labs(title = "Bookings by ADR Range", x = "ADR Bin", y = "Number of Bookings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),plot.title = element_text(hjust = 0.5))
```

This plot shows the distribution of bookings across different ADR (Average Daily Rate) ranges. Our goal was to identify the most commonly booked price ranges. The graph reveals that the majority of bookings fall between ADR values of 50 and 125. The distribution forms a bell-shaped curve, with a steeper decline on the lower ADR side and a more gradual drop-off as ADR increases. This suggests that while most guests book within a moderate price range, higher rates still receive a steady number of bookings, whereas lower rates are less frequent.

```{r, warning=FALSE, message=FALSE}
cleaned_data <- data %>%
  filter(!is.na(adr), !is.na(lead_time), adr <= 1000)

ggplot(cleaned_data, aes(x = lead_time, y = adr)) +
  geom_point(alpha = 0.15) +  
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  labs(
    title = "Relationship Between Lead Time and Average Daily Rate (ADR)",
    x = "Lead Time (Days)",
    y = "Average Daily Rate (ADR)"
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

This graph explores the relationship between lead time (how far in advance a hotel is booked) and the Average Daily Rate (ADR). Our prediction was that booking earlier would result in a lower ADR. The linear regression line shows a negative trend, generally supporting our hypothesis, that the earlier a booking is made, the lower the ADR tends to be. Most bookings occur with a lead time under 300 days, and while bookings made more than 300 days in advance are less common, they tend to have ADRs within a relatively narrow range. In contrast, shorter lead times show a wider spread of ADR values.



```{r, warning=FALSE, message=FALSE}
# ADR by month
monthly_adr_data <- data %>%
  group_by(arrival_date_month) %>%
  summarise(avg_adr = mean(adr, na.rm = TRUE), bookings = n())

# Average bookings by month
monthly_bookings <- data %>%
  group_by(arrival_date_month, arrival_date_year) %>%
  summarise(monthly_bookings = n()) %>%
  ungroup()

# bookings by month looking at years
monthly_avg_bookings <- monthly_bookings %>%
  group_by(arrival_date_month) %>%
  summarise(avg_bookings = mean(monthly_bookings, na.rm = TRUE)) 

# Create labeled datasets
monthly_avg_bookings$metric <- "Average Bookings"
monthly_avg_bookings$value <- monthly_avg_bookings$avg_bookings

monthly_adr_data$metric <- "Average ADR"
monthly_adr_data$value <- monthly_adr_data$avg_adr

# Combine sets into data frame
combined_monthly_data <- rbind(
  monthly_avg_bookings[, c("arrival_date_month", "metric", "value")],
  monthly_adr_data[, c("arrival_date_month", "metric", "value")]
)

# Set month order
combined_monthly_data$arrival_date_month <- factor(
  combined_monthly_data$arrival_date_month,
  levels = levels(data$arrival_date_month)
)

# display both plots together
ggplot(combined_monthly_data, aes(x = arrival_date_month, y = value, fill = arrival_date_month)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_manual(values = monthcolors) +
  facet_wrap(~metric, scales = "free_y") +  # makes it side-by-side for comparison, personal preference
  labs(title = "Monthly Comparison of Average Bookings and ADR",
       x = "Month", y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),plot.title = element_text(hjust = 0.5))

```

These two graphs were created to analyze the differences in bookings and ADR (Average Daily Rate) throughout the year. Our initial prediction was that booking volume would increase in the spring, peak during the summer, decline in the fall, and be lowest in the winter. Based on this expectation, we also assumed that ADR would follow a similar trend—rising during high-demand months when more people tend to go on vacation. While the ADR trend generally aligned with our prediction, peaking in the summer months, the bookings data did not. In July and August, months we expected to have the highest volume of bookings, we observed a significant drop. This contrast suggests that although prices were higher in those months, it did not directly correlate with an increase in bookings as we anticipated.


```{r, message=FALSE, warning=FALSE}
# clean data and hide extreme outliers
filtered_data <- data %>%
  mutate(total_guests = adults + children + babies) %>%
  filter(adr <= 1000, total_guests > 0, total_guests < 20)

ggplot(filtered_data, aes(x = total_guests, y = adr)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "ADR vs. Number of Guests",
       x = "Total Number of Guests",
       y = "Average Daily Rate (ADR)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

The purpose of this graph is to analyze how the Average Daily Rate (ADR) changes as the number of guests in a booking increases. For guest counts up to about 5, the ADR values appear to have a consistent range. However, as the number of guests increases beyond that, especially in higher guest group sizes, the ADR tends to vary more and generally leans lower, as indicated by the outliers. The linear regression line shows an upward trend, implying that ADR tends to increase with more guests. Still, this relationship is not strict because there are many exceptions, and the data shows higher variability at higher guest counts.


```{r, message=FALSE, warning=FALSE}

data <- data %>%
  mutate(has_agent = ifelse(is.na(agent), "No Agent", "Has Agent"))

# Filter for ADR values below 1000
filtered_data <- data %>%
  filter(adr < 1000)

# Box plot of ADR by agent status
ggplot(filtered_data, aes(x = has_agent, y = adr, fill = has_agent)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(
    title = "ADR Comparison (ADR < 1000): Bookings With vs Without Agent",
    x = "Agent Status",
    y = "Average Daily Rate (ADR)"
  ) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

```

This graph was created to explore how booking through an agent or travel agency affects the Average Daily Rate (ADR). Our initial assumption was that using an agent would result in a lower ADR, possibly due to negotiated discounts, package deals, or just having more experience. However, the data showed the opposite: bookings made through an agent tended to have a slightly higher ADR across all four quartiles. Additionally, agent bookings displayed a larger number of outliers, indicating greater variability in prices compared to direct bookings.
  
  

### How does number of guests affect hotel bookings?


```{r, warning=FALSE, message=FALSE}

# find total number of guests
data <- data %>%
  mutate(number_of_guests = adults + children + babies)

guest_counts <- data %>%
  group_by(number_of_guests) %>%
  summarise(booking_count = n()) %>%
  ungroup()

# Plot 
ggplot(guest_counts, aes(x = factor(number_of_guests), y = booking_count)) +
  geom_bar(stat = "identity", fill="blue") +
  geom_text(aes(label = booking_count), vjust = -0.5, size = 3) +
  labs(
    title = "Number of Bookings by Guest Count",
    x = "Number of Guests",
    y = "Number of Bookings"
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

This graph illustrates the frequency of bookings based on the number of guests. The most common group size by far was 2 guests. Bookings with only 1 guest occurred significantly less, nearly four times less frequent. The frequency drops sharply for group sizes larger than 2, with bookings for more than 5 guests being relatively rare and appearing as nearly isolated cases.



```{r, warning=FALSE, message=FALSE}

# Find total number of guests
data <- data %>%
  mutate(
    number_of_guests = adults + children + babies
  ) %>%
  filter(!is.na(number_of_guests), number_of_guests > 0)  

# Find cancellation rate by group size
group_cancellation <- data %>%
  group_by(number_of_guests) %>%
  summarise(
    total = n(),
    cancelled = sum(is_canceled),
    cancellation_rate = cancelled / total
  )

ggplot(group_cancellation, aes(x = factor(number_of_guests), y = cancellation_rate)) +
  geom_col(fill = "orange") +
  labs(title = "Cancellation Rate by Group Size", x = "Group Size (Number of Guests)", y = "Cancellation Rate") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5))

```

The purpose of this graph was to examine how group size affects the cancellation rate. We hypothesized that larger groups would have higher cancellation rates due to the difficulty of coordinating multiple schedules. The data shows the following: bookings for groups of 5 or fewer had cancellation rates between 25% and 40%, while bookings for groups larger than 5 were canceled at least 50% of the time, and more commonly 100% of the time.




```{r, warning=FALSE, message=FALSE}

# Find average lead time and ADR by group size
group_behavior <- data %>%
  group_by(number_of_guests) %>%
  summarise(
    avg_lead_time = mean(lead_time, na.rm = TRUE),
    avg_adr = mean(adr, na.rm = TRUE)
  )

# Plot
ggplot(group_behavior, aes(x = factor(number_of_guests), y = avg_lead_time)) +
  geom_col(fill = "purple") +
  labs(title = "Average Lead Time by Group Size", x = "Group Size", y = "Average Lead Time (days)") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5))

```

This graph was created to analyze the average lead time based on group size. We expected that larger groups would book further in advance to allow more time for coordination and planning. Interestingly, for smaller group sizes (up to 5), the trend was the opposit where average lead time decreased slightly as group size increased, with groups of 2 booking the furthest in advance at just over 120 days. However, for groups larger than 5, the trend reversed: most of these group sizes had average lead times exceeding 300 days, with the exceptions being groups of 10 and 12.




```{r, warning=FALSE, message=FALSE}

data_non_canceled <- data %>%
  filter(is_canceled == 0)

# Group by guest count and parking spots and then find count frequency
data_grouped <- data_non_canceled %>%
  group_by(number_of_guests, required_car_parking_spaces) %>%
  summarise(count = n(), .groups = 'drop')

# Plot with point size representing frequency
ggplot(data_grouped, aes(x = number_of_guests, y = required_car_parking_spaces, size = count)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", color = "orange", se = FALSE, show.legend = FALSE) +
  scale_size_continuous(range = c(1, 8)) +
  labs(
    title = "Number of Guests vs Car Parking Spots",
    subtitle = "Point size reflects frequency of booking combinations",
    x = "Number of Guests",
    y = "Required Car Parking Spots",
    size = "Booking Count"
  ) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))

```

The purpose of this graph is to show the relationship between the number of required parking spots and the number of guests in a booking. We hypothesized that as the number of guests increased, the average number of required parking spots would also increase. To ensure accuracy, the data was filtered to exclude canceled bookings. However, our prediction was proven wrong by the results. The linear regression line shows a slight negative trend, indicating that higher guest counts do not correlate with more parking spots. In fact, we observed that most bookings required no parking spots at all, regardless of guest count. When parking was requested, it was typically limited to just one spot.




```{r, warning=FALSE, message=FALSE}

# Summarize count of bookings and average parking spots per number of guests
guest_parking_summary <- data_non_canceled %>%
  group_by(number_of_guests) %>%
  summarise(
    booking_count = n(),
    avg_parking_spots = mean(required_car_parking_spaces, na.rm = TRUE)
  ) %>%
  filter(number_of_guests <= 20) 

# Plot
ggplot(guest_parking_summary, aes(x = factor(number_of_guests), y = booking_count)) +
  geom_bar(stat = "identity", fill = "blue") +
  geom_text(aes(label = paste0("Avg Spots: ", round(avg_parking_spots, 3))), 
            vjust = -0.5, size = 3) +
  labs(
    title = "Booking Count by Guest Group Size",
    subtitle = "Labels above bars show average required car parking spots",
    x = "Number of Guests",
    y = "Number of Bookings"
  ) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.subtitle=element_text(hjust=0.5))

```

This graph again examines the relationship between the number of guests and the required parking spots, but this time it highlights the average number of parking spots for each guest count. From this visualization, we can see that the average number of parking spots generally increases with group size up to around four guests. After that point, the average begins to decline, and for group sizes larger than five, the need for parking spots drops significantly, with many of those bookings requiring none at all.




```{r, warning=FALSE, message=FALSE} 

data <- data %>%
  mutate(
    number_of_guests = adults + children + babies,
    total_nights = stays_in_weekend_nights + stays_in_week_nights
  )

# Filter for bookings where guests showed
filtered_data <- data %>%
  filter(number_of_guests > 0, number_of_guests <= 20, total_nights > 0)

# Plot
ggplot(filtered_data, aes(x = number_of_guests, y = total_nights)) +
  geom_jitter(alpha = 0.2, color = "blue", width = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  labs(
    title = "Number of Guests vs Nights Stayed",
    x = "Number of Guests",
    y = "Total Nights Stayed"
  ) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5))

```

This graph was created to explore how the number of guests affects the number of nights stayed per booking. The linear regression line shows a slight upward trend, suggesting that bookings with more guests may stay longer on average. However, it's important to note that most bookings were made for one or two guests, with bookings for two guests being nearly four times more common than those for one. This imbalance in the data may skew the regression line. In fact, when looking more closely at the scatter, we can see a subtle downward trend in nights stayed as guest count increases beyond two. This suggests that while larger groups do occasionally stay longer, they may also book shorter and frequent stays, making it ore difficult to understand the relationship between guest count and duration.

  
  
  
### What affects how many times bookings are changed?  

We are going to examine the relationship between number of booking changes made with:  
- cancellation rate  
- hotel type  
- lead time  
- returning guest status  
- customer type  
- deposit type  

```{r, message=FALSE, warning=FALSE}

data %>%
  count(booking_changes) %>%
  filter(n > 50) %>%
  ggplot(aes(x = as.factor(booking_changes), y = n)) +
  geom_bar(stat = "identity", fill = "pink") +
  labs(
    title = "Distribution of Booking Changes",
    x = "Number of Changes",
    y = "Number of Bookings"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5)
  )


```
  
This chart shows the distribution of how many booking changes are made. As you can see, the majority of bookings don't have any changes made to them, and it drops from there.  


This is the cancellation rate by number of booking changes.  

```{r, message=FALSE, warning=FALSE}


data %>%
  group_by(booking_changes) %>%
  summarise(
    total = n(),
    cancel_rate = mean(is_canceled),
    .groups = "drop"
  ) %>%
  filter(total > 10) %>%
  ggplot(aes(x = as.factor(booking_changes), y = cancel_rate)) +
  geom_bar(stat = "identity", fill = "orange") +
  geom_text(aes(label = scales::percent(cancel_rate, accuracy = 0.1)), 
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Cancellation Rate by Number of Booking Changes",
    x = "Number of Changes",
    y = "Cancellation Rate"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5)
  )


```
  
This shows cancellation rate by number of booking changes. As shown, there doesn't seem to be much correlation between the number of booking changes and the cancellation rate. 0 booking changes has the highest rate of cancellation but 0 booking changes also has the highest number of occurrences out of the booking changes. Overall, the data from this graph isn't significantly helpful in our analysis, but we originally thought that bookings with more changes made would be significantly higher for cancellation rates because in theory that customer would have plans that were not solid.  


This shows booking changes by hotel type.  
```{r, message=FALSE, warning=FALSE}

ggplot(data, aes(x = hotel, y = booking_changes, fill = hotel)) +
  stat_summary(fun = mean, geom = "bar", width = 0.6, show.legend = FALSE) +
  labs(
    title = "Average Number of Booking Changes by Hotel Type",
    x = "Hotel Type",
    y = "Average Booking Changes"
  ) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5))

```
  
As we can see the hotel type doesn't seem to affect the number of booking changes a lot, but the resort hotel does have a slightly higher average number of booking changes at about 0.3, whereas the city hotel has an average closer to 0.2 booking changes made. This is still relatively similar though. This could potentially be because resort hotels are more vacation-based so they probably have more groups going for fun that want booking changes to be made for convenience, as they don't have to plan as strictly as groups like coworkers or teams that might book more frequently in cities.    

How does lead time correlate to booking changes? We hypothesize the longer lead time is the more booking changes will be made.   
```{r, message=FALSE, warning=FALSE}


data %>%
  mutate(lead_time_bin = cut(lead_time, breaks = 20)) %>%  # this is like a bin, just had to do it a different way than normal
  group_by(lead_time_bin) %>%
  summarise(avg_booking_changes = mean(booking_changes), .groups = 'drop') %>%
  ggplot(aes(x = lead_time_bin, y = avg_booking_changes)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Booking Changes by Lead Time",
    x = "Lead Time",
    y = "Average Number of Booking Changes"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )


```
  
As shown in the graph there doesn't seem to be too much change in booking changes with lead time. It stays less than 0.5 average booking changes until we get to the 700 to 738 day lead time range, then it significantly spikes and the average number of booking changes is 3. However, there are very few values in that range, so likely it was just one booking that was changed 3 times.  

The following graph demonstrates the relationship between whether a customer is a returning guest and the number of booking changes made.  
```{r, message=FALSE, warning=FALSE}

  ggplot(data, aes(x = is_repeated_guest, y = booking_changes, fill = is_repeated_guest)) +
  stat_summary(fun = mean, geom = "bar", width = 0.6) +
  labs(
    title = "Average Number of Booking Changes by Returning Guest",
    x = "Guest Status",
    y = "Average Booking Changes"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )

```
  
As shown in the graph we see that repeated guests have a slightly higher chance of changing their bookings. However, both have a small average number of booking changes with repeating guests being around 0.26 changes made on average and new guests being around 0.22 changes made on average.  


The following graph shows customer type vs number of booking changes made.  
```{r, message=FALSE, warning=FALSE}


# this reorders so that they go lowest to highest average num of booking changes
data <- data %>%
  group_by(customer_type) %>%
  mutate(avg_changes = mean(booking_changes, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(customer_type = reorder(customer_type, avg_changes))


ggplot(data, aes(x = customer_type, y = booking_changes, fill = customer_type)) +
  stat_summary(fun = mean, geom = "bar", width = 0.6) +
  labs(
    title = "Average Number of Booking Changes by Customer Type",
    x = "Customer Type",
    y = "Average Booking Changes"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )


```  
  
This graph shows that transient-party customer types are more likely to make more booking changes than the others with an average number of booking changes of about 0.35, and contract customers make the least amount of booking changes with an average around 0.11. This makes sense as with more people in a party there is bound to be more changes that have to be made to accommodate for more schedules, and transient groups have the most flexibility.    
  
Now we investigate the relationship between deposit type and number of booking changes.  
```{r, message=FALSE, warning=FALSE}

data %>%
  group_by(deposit_type) %>%
  summarise(
    avg_booking_changes = mean(booking_changes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(deposit_type = reorder(deposit_type, avg_booking_changes)) %>%
  ggplot(aes(x = deposit_type, y = avg_booking_changes, fill = deposit_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(avg_booking_changes, 2)), 
            vjust = -0.5, size = 4) +
  labs(
    title = "Average Number of Booking Changes by Deposit Type",
    x = "Deposit Type",
    y = "Average Booking Changes"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )


```
  
As you can see, refundable deposit types have a higher average number of changes made to the bookings at 0.59 on average, whereas no deposit is next at 0.25, and non refund has a very low number of changes on average at 0.01. This agrees with our original hypothesis, as refundable deposit types allow customers more booking freedom which would allow for them to make changes easier.   
  
### Where are most visitors coming from?   
  
To answer this question we observe the country of origin of bookings made.  
```{r, message=FALSE, warning=FALSE}


country_counts <- data %>%
  group_by(country) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  top_n(10, count)  # limit to top 10 countries or else the graph is too big 

ggplot(country_counts, aes(x = reorder(country, -count), y = count, fill = country)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Top Countries by Number of Hotel Bookings",
    x = "Country",
    y = "Number of Bookings"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )



```
  
This information would be useful for these hotels to know which countries to focus their advertising on. As shown in the graph, the top 10 countries for number of bookings made is Portugal, the UK, France, Spain, Germany, Italy, Ireland, Belgium, Brazil, and the Netherlands. However, Portugal by FAR has the most bookings with almost 5,000, and the next largest is only around 1200 to 1300 bookings.  
  
### Conclusion  
  
  



