---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Research topic: Hotel Booking Data 

### Team members: Kaitlyn Hoyme and Christopher Moseley  

### Introduction:  

This report will demonstrate various relationships between variables surrounding hotel booking data. We will examine these variables in order to determine how they affect each other and hotel bookings. This data would be of interest to businesses like hotels, as they can adjust these factors to improve their profit and customer satisfaction.  

  
### Questions we will address: 

- How does the time of year affect hotel bookings?  
- What makes guests more likely to book these hotels / cancel their bookings?  
- How does the average daily rate affect bookings?  
- How does number of guests affect hotel bookings?  
- How does hotel type impact bookings?  
  
By investigating these questions we hope to demonstrate useful data about what factors affect hotel bookings.    
  

### Data:   

This dataset includes information on hotel booking data from two hotels, one a city hotel, the other a resort hotel, with data collected between the 1st of July 2015 and 31st of August 2017.   

There are 119390 rows of 36 variables.  
The Variables are:   
 - hotel : States what hotel the row represents data from. City Hotel or Resort Hotel.  
 - is_canceled : value that indicates if the booking was canceled (1) or not (0).  
 - lead_time : Number of days that occurred between the entering date of the booking into the PMS and the arrival date.  
 - arrival_date_year : The year of the arrival date (values are between 2015-2017).  
 - arrival_date_month : Month of arrival date. Categorical variable with values "January" through "December".  
 - arrival_date_week_number : Week number of the arrival date.  
 - arrival_date_day_of_month : Day of the month of the arrival date.  
 - stays_in_weekend_nights : Number of weekend nights (Saturday or Sunday) the guest stayed or booked to stay at the hotel.   
 - stays_in_week_nights :  Number of week nights (Monday-Friday) the guest stayed or booked to stay at the hotel.   
 - adults : Number of adults.  
 - children : Number of children.  
 - babies : Number of babies.  
 - meal : Represents the type of meals offered by the hotel. Ex: BB = Bed & Breakfast.  
 - country : Country of origin.  
 - market_segment : Market segment designation. In categories (categorical). "TA" means travel agents, and "TO" means tour operators    
 - distribution_channel : Booking distribution channel. The term "TA" means travel agents, and "TO" means tour operators.  
 - is_repeated_guest : Value indicating if the booking name was from a repeated guest (1) or not (0).  
 - previous_cancellations : Number of previous bookings that were cancelled by the customer prior to the current booking.   
 - previous_bookings_not_canceled : Number of previous bookings not cancelled by the customer prior to the current booking.   
 - reserved_room_type : Code of room type reserved.   
 - assigned_room_type : Code for type of room assigned to the booking. Sometimes the assigned room type differs from the reserved room type due to hotel operation reasons (eg overbooking) or by customer request.     
 - booking_changes : Number of changes/amendments made to the booking from the moment the booking was entered on the PMS until the moment of check-in or cancellation.   
 - deposit_type : Represents deposit type. No Deposit - no deposit was made; Non Refund - a deposit was made in the value of the total stay cost; Refundable - a deposit was made with a value under the total cost of stay.   
 - agent : ID of the travel agency that made the booking.  
 - company : ID of the company/entity that made the booking or responsible for paying the booking. ID is presented instead of designation for anonymity reasons.    
 - days_in_waiting_list : Number of days the booking was in the waiting list before it was confirmed to the customer.   
 - customer_type : Group - when the booking is associated to a group; Transient - when the booking is not part of a group or contract and is not associated to other transient booking; Transient-party - when the booking is transient but is associated to at least one other transient booking.   
 - adr : Average daily rate calculated by dividing the sum of all lodging transactions by the total number of staying nights.  
 - required_car_parking_spaces : Number of parking spaces required by the customer.   
 - total_of_special_requests : Number of special requests made by the customer (e.g. twin bed or high floor).  
 - reservation_status : Check-Out - customer has checked in but already departed; No-Show - customer did not check-in and did inform the hotel of the reason why, etc..  
 - reservation_status_date : Date at which the last status was set. This variable can be used in conjunction with the ReservationStatus to understand when the booking was canceled or when the customer check-out of the hotel.  
 - name : Name of the guest (not real for anonymity).  
 - email : Email of the guest (not real for anonymity).  
 - phone-number : Phone number of the guest (not real for anonymity).  
 - credit_card : Credit Card Number of the guest (not real for anonymity).    
 
This dataset was found through Kaggle and all of the data recorded is real. However, for reasons surrounding anonymity variables like name, email, phone-number, and credit card are all made up. The data was originally obtained from "Hotel Booking Demand Datasets", an article written by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.   
This is a link to our dataset on Kaggle: https://www.kaggle.com/datasets/mojtaba142/hotel-booking/data  

### Data Cleaning
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
#Data: description of your data set, first data cleaning steps, marginal summaries;

url <- "https://raw.githubusercontent.com/khoy24/DS202FinalProject/main/hotel_booking.csv"
data <- read.csv(url)
colnames(data)


nrow(data) # 119390 rows/entries
ncol(data) # 36 variables

any(is.na(data))
# no missing entries

duplicateRows <- data %>% filter(duplicated(.))
print(nrow(duplicateRows))
#no duplicate rows

#examine types of columns, make sure makes sense
str(data)





```

### Results
  
  
### Main - Curiosity  
Intense exploration and evidence of many trials and failures. Presents best ideas, rather than all ideas. Additional research from other sources used to help understand/explain findings.  


How does the time of year affect hotel bookings?  
```{r, message=FALSE, warning=FALSE}

#This plot shows the number of hotel bookings per month (it shows the number of bookings made FOR that month, not what month the booking was made).

#ended up having to make months a factor to reorder in order of correct months
data$arrival_date_month <- factor(data$arrival_date_month, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

ggplot(data, aes(x = arrival_date_month, fill=arrival_date_month)) +
  geom_bar(width = 0.8) + 
  labs(title = "Hotel Bookings per Month", x = "Month", y = "Number of Bookings") +
  #can only have one theme() call in a ggplot otherwise it gets overwritten, so I combined the qualities into one
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.title = element_text(hjust = 0.5) #needed this to center the title 
  )

#As we can see, number of bookings are much higher in the summer months, and much lower in the winter months.


#Now we check what year had more bookings made

#After making the first graph realized we needed to make year not be an int. Change that and the similar values to strings.
data <- data %>%
  mutate(
    `arrival_date_year` = as.character(`arrival_date_year`),
    `arrival_date_week_number` = as.character(`arrival_date_week_number`),
    `arrival_date_day_of_month` = as.character(`arrival_date_day_of_month`),
  )


ggplot(data, aes(x = arrival_date_year, fill=arrival_date_year)) +
  geom_bar(width = 0.8) + 
  labs(title = "Hotel Bookings by Year", x = "Year", y = "Number of Bookings") +
  #can only have one theme() call in a ggplot otherwise it gets overwritten, so I combined the qualities into one
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.title = element_text(hjust = 0.5) #needed this to center the title 
  )

# As we can see, 2015 had little bookings, but we also only have data for half the year of 2015, and 8 months of 2017. 2016 was the only full year we had. Because of this maybe there is a better way to represent this, but it probably isn't as good as a representative of the data as the months. 

#Might also want to separate hotel bookings per month and facet that by year, as we'd have more months for those included on the half years. 

ggplot(data, aes(x = arrival_date_month, fill = arrival_date_month)) +
  geom_bar(width = 0.8) + 
  labs(title = "Hotel Bookings per Month (Faceted by Year)", 
       x = "Month", y = "Number of Bookings") +
  facet_wrap(~ arrival_date_year) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  )

#Now we see that in 2015 the months with the most bookings were September-October, in 2016 it was October again, followed by the spring and summer months, and in 2017 it was May with the highest number of bookings, with its surrounding months following it. For the most part we still see that months with warmer weather have more bookings.  
```
