---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Research topic: Hotel Booking Data 

### Team members: Kaitlyn Hoyme and Christopher Moseley  

### Introduction:  

This report will demonstrate various relationships between variables surrounding hotel booking data. We will examine these variables in order to determine how they affect hotel bookings. This data would be of interest to businesses like hotels, as they can adjust these factors to improve profit and customer satisfaction.  

### Data:   

This dataset includes information on hotel booking data from two hotels, one a city hotel, the other a resort hotel, between the 1st of July 2015 and 31st of August 2017.  

 There are 119390 rows of 36 variables.  
 Variables are:  
 - hotel : States what hotel the row represents data from. City Hotel or Resort Hotel.  
 - is_canceled : value that indicates if the booking was canceled (1) or not (0).  
 - lead_time : Number of days that occurred between the entering date of the booking into the PMS and the arrival date.  
 - arrival_date_year : The year of the arrival date (values are between 2015-2017).  
 - arrival_date_month : Month of arrival date. Categorical variable with values "January" through "December".  
 - arrival_date_week_number : Week number of the arrival date.  
 - arrival_date_day_of_month : Day of the month of the arrival date.  
 - stays_in_weekend_nights : Number of weekend nights (Saturday or Sunday) the guest stayed or booked to stay at the hotel.   
 - stays_in_week_nights :  Number of week nights (Monday-Friday) the guest stayed or booked to stay at the hotel.   
 - adults : Number of adults.  
 - children : Number of children.  
 - babies : Number of babies.  
 - meal : Represents the type of meals offered by the hotel. Ex: BB = Bed & Breakfast.  
 - country : Country of origin.  
 - market_segment : Market segment designation. In categories (categorical). "TA" means travel agents, and "TO" means tour operators    
 - distribution_channel : Booking distribution channel. The term "TA" means travel agents, and "TO" means tour operators.  
 - is_repeated_guest : Value indicating if the booking name was from a repeated guest (1) or not (0).  
 - previous_cancellations : Number of previous bookings that were cancelled by the customer prior to the current booking.   
 - previous_bookings_not_canceled : Number of previous bookings not cancelled by the customer prior to the current booking.   
 - reserved_room_type : Code of room type reserved.   
 - assigned_room_type : Code for type of room assigned to the booking. Sometimes the assigned room type differs from the reserved room type due to hotel operation reasons (eg overbooking) or by customer request.     
 - booking_changes : Number of changes/amendments made to the booking from the moment the booking was entered on the PMS until the moment of check-in or cancellation.   
 - deposit_type : Represents deposit type. No Deposit - no deposit was made; Non Refund - a deposit was made in the value of the total stay cost; Refundable - a deposit was made with a value under the total cost of stay.   
 - agent : ID of the travel agency that made the booking.  
 - company : ID of the company/entity that made the booking or responsible for paying the booking. ID is presented instead of designation for anonymity reasons.    
 - days_in_waiting_list : Number of days the booking was in the waiting list before it was confirmed to the customer.   
 - customer_type : Group - when the booking is associated to a group; Transient - when the booking is not part of a group or contract and is not associated to other transient booking; Transient-party - when the booking is transient but is associated to at least one other transient booking.   
 - adr : Average daily rate calculated by dividing the sum of all lodging transactions by the total number of staying nights.  
 - required_car_parking_spaces : Number of parking spaces required by the customer.   
 - total_of_special_requests : Number of special requests made by the customer (e.g. twin bed or high floor).  
 - reservation_status : Check-Out - customer has checked in but already departed; No-Show - customer did not check-in and did inform the hotel of the reason why, etc..  
 - reservation_status_date : Date at which the last status was set. This variable can be used in conjunction with the ReservationStatus to understand when the booking was canceled or when the customer check-out of the hotel.  
 - name : Name of the guest (not real for anonymity).  
 - email : Email of the guest (not real for anonymity).  
 - phone-number : Phone number of the guest (not real for anonymity).  
 - credit_card : Credit Card Number of the guest (not real for anonymity).    
 
 
 This dataset was found through Kaggle, the data is real. The data is originally from "Hotel Booking Demand Datasets", an article written by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.   


```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
#Data: description of your data set, first data cleaning steps, marginal summaries;

url <- "https://raw.githubusercontent.com/khoy24/DS202FinalProject/main/hotel_booking.csv"
data <- read.csv(url)
colnames(data)


nrow(data) # 119390 rows/entries
ncol(data) # 36 variables

any(is.na(data))
# no missing entries

duplicateRows <- data %>% filter(duplicated(.))
print(nrow(duplicateRows))
#no duplicate rows



calculate_summary <- function(column) {
  c(
    mean = mean(column, na.rm = TRUE),
    median = median(column, na.rm = TRUE),
    sd = sd(column, na.rm = TRUE),
    min = min(column, na.rm = TRUE),
    max = max(column, na.rm = TRUE),
    range = diff(range(column, na.rm = TRUE))  # Range is the difference between max and min
  )
}

#will have to redo this for new dataset

#summary_stats <- data.frame(
#  age = calculate_summary(data$age),
#  income = calculate_summary(data$income),
#  purchase_amount = calculate_summary(data$purchase_amount)
#)

#print(summary_stats)



```

  
### Questions to be addressed: 

- How does the time of year affect hotel bookings?  
- What makes guests more likely to book these hotels / cancel their bookings?  
- How does the average daily rate affect bookings?  
- How does number of guests affect hotel bookings?  
- How does hotel type impact bookings?  

  
  
### Main - Curiosity  
Intense exploration and evidence of many trials and failures. Presents best ideas, rather than all ideas. Additional research from other sources used to help understand/explain findings.  


```{r, message=FALSE, warning=FALSE}

#practice plots

ggplot(data, aes(x = hotel, fill = factor(is_canceled))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Cancellations by Hotel Type",
       x = "Hotel Type", y = "Proportion",
       fill = "Cancellation") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

ggplot(data, aes(x = adults + children + babies, fill = factor(is_canceled))) +
  geom_bar(position = "fill") +
  labs(title = "Booking Size vs Cancellations",
       x = "Number of People (Adults + Children + Babies)", y = "Proportion",
       fill = "Cancellation") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

ggplot(data, aes(x = is_repeated_guest, fill = factor(is_canceled))) +
  geom_bar(position = "fill") +
  labs(title = "Cancellation Rate Among Repeat Guests",
       x = "Repeat Guest", y = "Proportion",
       fill = "Cancellation") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

ggplot(data, aes(x = meal, y = adr, fill = meal)) +
  geom_boxplot() +
  labs(title = "Meal Type vs ADR",
       x = "Meal Type", y = "ADR (Average Daily Rate)") +
  theme_minimal()

ggplot(data, aes(x = lead_time, fill = factor(is_canceled))) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  labs(title = "Cancellation by Lead Time",
       x = "Lead Time (days)", y = "Count",
       fill = "Cancellation") +
  theme_minimal()

max(data$adults)
max(data$babies)
max(data$children)


```
