---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Research topic: Customer Purchases Behavior  

### Team members: Kaitlyn Hoyme and Christopher Moseley  

### Data:   

This dataset includes information on customer purchasing behavior.  
 There are 100000 rows of 12 variables.  
 Variables are:  
 - age : age of customer   
 - gender : gender of customer  
 - income : income of customer  
 - education : education level of customer  
 - region : region of customers residence  
 - loyalty_status : loyalty status of customer  
 - purchasing_frequency : frequency of purchases by customer  
 - purchase_amount : amount spent by customer in each purchase  
 - product_category : category of purchased product  
 - promotion_usage : 0 means the customer did not use promotional offers, 1 means the customer did use promotional offers  
 - satisfaction_score : satisfaction score of customer  
 
 This data was originally simulated in R and the dataset was obtained through Kaggle, the data mimics real scenarios.  


```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

#Data: description of your data set, first data cleaning steps, marginal summaries;

url <- "https://raw.githubusercontent.com/khoy24/DS202FinalProject/main/customer_data.csv"
data <- read.csv(url)
 
any(is.na(data))
# no missing entries

duplicateRows <- data %>% filter(duplicated(.))
print(nrow(duplicateRows))
#no duplicate rows

```

  
### Questions to be addressed: 

- How does gender affect customer purchasing behavior?
 - How does region affect customer purchasing behavior?
 - How does education level affect customer purchasing behavior?
 - How does income affect customer purchasing behavior?
 - How does age affect customer purchasing behavior?
 - What category of products are more popular among various age groups, genders, and regions?
 - How does loyalty status of the customer affect purchasing behavior?
 - How do various factors (age,gender,region,education level,income) affect whether or not a customer has a loyalty status?
 - What is the relationship between purchase frequency and purchase amount?
 - How does promotion usage affect customer behavior?
 - How do various factors (age,gender,region,education level,income) affect customer satisfaction score?
  
